\textbf{ Matching the forecast horizon with the relevant ecological processes }

### 2019-09-16 doc/manuscript.tex
Addressed Morales' comments

\author[1]{Peter B. Adler}  %\thanks{Corresponding author. Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah Email: peter.adler@usu.edu}}
\author[2]{Ethan P. White}
\author[3]{Michael H. Cortez}
\affil[1]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah}
\affil[2]{Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida}
\affil[3]{Department of Biological Science, Florida State University, Tallahasee, Florida}

@@

We made forecasts based on two phenomenological models, each representing processes operating at different time scales. 
The first phenomenological model represents the space-for-time substiution approach, which we call the ``spatial approach'' for brevity. 
We correlated the mean environment with the mean of an ecological state or rate across many sites. The second model represents the time-series or ``temporal approach.'' We correlated interannual variation in an ecological response with interannual variation in the environment at just one site. 
We compared forecasts from both models to the simulated dynamics to determine how well the two approaches perform at different forecast horizons. We also assessed the potential for combining the information available in both spatial and temporal patterns by using a weighted average of the forecasts from the spatial and temporal approaches optimized to best match the (simulated) observations. We then studied how the optimal model weights changed over time. These spatial, temporal and combined statistical
models are described in Appendix \ref{models}. We expected the temporal model to best predict short-term dynamics, the spatial model to best predict long-term dynamics, and that a combination of the two approaches would provide the best forecasts at transitional, intermediate time scales.

@@

Conservation biologists and natural resource managers often need to anticipate the impact of environmental change on the abundance of 
endangered species, biological invaders, and harvested species. Although the managers may be primarily interested in just one focal species, skillful prediction might require considering interactions with many other species, greatly complicating the problem. But at what forecast horizon do altered species interactions become impossible to ignore? We explored this question using a metacommunity model developed by \cite{alexander_lags_2018}. The model describes Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient. Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by  fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability. We provide a more detailed description of the model in Appendix \ref{metacomm}.

@@

During the baseline period there were strong spatial patterns across the mean temperature gradient. 
Individual species, including our focal species, showed classic, unimodal ``Whittaker" patterns of abundances across the gradient (Fig. \ref{fig:species-patterns-models}A).
These spatial patterns are the basis for our ``spatial model" of the temperature-biomass relationship for our focal species (Fig. \ref{fig:species-patterns-models}A). 
In contrast to the strong spatial patterns, population and community responses to interannual variation in temperature within sites were weak.
At our focal site in the center of the gradient, the biomass of the focal species was quite insensitive to interannnual variation in temperature (Fig. \ref{fig:species-patterns-models}B). Our ``temporal model" estimates this weak, linear temperature effect, along with strong lag effects of biomass in the previous year.

@@

We forecast the effect of a temperature increase (Fig. \ref{fig:community-forecast-species}A) on the focal species' biomass at one location in the center of the temperature gradient 
using both the spatial model and the temporal model. 
The predictions from the spatial and temporal models contrasted markedly, with the temporal 
model predicting a large increase in biomass and the spatial model predicting a decrease. Initially, the simulated abundances followed the increase predicted by the 
temporal model, but as faster-growing species colonized and increased in abundance at the focal site, the biomass of the focal species decreased, eventually falling below its baseline level (Fig. \ref{fig:community-forecast-species}B).

To combine the spatial and temporal model into a single forecast, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, the combined population forecast is $\omega \cdot T(N_{t-1},K_t) + (1-\omega) \cdot S(K_t) $ where $T$ is the temporal model, which depends on population size, $N$, and expected temperature, $K$, and $S$ is the spatial model, which depends only on $K$ (see Appendix \ref{models} for full description of the approach).
This combined model accurately predicts the simulated dynamics across the full forecast horizon (Fig. \ref{fig:community-forecast-species}B), but 
the reason it works so well is that the weights were determined by fitting directly to the data. Unlike our spatial and temporal model forecasts, 
we did not generate out-of-sample predictions from the combined model; it merely provides 
a convenient way to quantify how rapidly dynamics shift from being dominated by interannual 
variation captured in the temporal model (time $t=0$ to $t \approx 1250$ in Fig. \ref{fig:community-forecast-species}B) to being dominated by the steady-state equilibrium captured by the spatial model (time $t \geq 2500$).
A true, combined forecast would require a method to determine the model weights \emph{a priori}.
The most rapid shifts in the model weights occurred during the period when warm-adapted, faster growing species were increasing most rapidly
in abundance (Fig. \ref{fig:community-forecast-species}C).

@@

The compositional turnover affecting our focal species also influences total biomass, linking community and ecosystem dynamics. We repeated our focal species analysis for total community biomass, and the results were similar: the temporal model initially made the best forecasts immediately following the onset of the temperature increase, but as the identity and abundances of species at the study site changed, the model weights rapidly shifted to the spatial model (Figs. \ref{fig:community-models-total} and \ref{fig:community-forecasts-total}).

@@

Evolutionary adaptation is a key uncertainty in predicting how environmental change will impact a focal population at a given location \citep{Hoffmann2011}. Like the shifts
in species composition illustrated in the previous example, shifts in genotype frequencies can also influence dynamics and forecasts at different time scales.
We demonstrate this process with a hypothetical annual plant population in which fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A). Our model describes how the local density of each genotype changes between years,  which depends on temperature and genotype densities in the previous year. Transient temporal dynamics are computed directly from the model; these dynamics define our temporal forecast. To create ``spatial data'', we simulated the equilibrium density of each genotype under different mean temperatures. The pattern of equilibrium densities across a gradient in mean annual temperature defines our spatial forecast: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B). The full model description is provided in Appendix \ref{eco-evo}.

@@

The change in model performance with increasing forecast horizon reflects differences in the types and scales of processes captured by spatial and temporal data sets. Local time-series data capture demographic processes, lagged effects, and responses of current assemblages to small changes in environmental conditions. In addition, the state of the system in the near future depends heavily on the current state. Since short-term forecasts do not typically require extrapolating into novel conditions, a model based on the historical range of variation which incorporates lags and accurate initial conditions is likely to be the most generally useful approach for forecasting. The hypothesis that time-series models will be most effective for near-term forecasts is supported by recent empirical analyses of biodiversity forecasts at time scales from one to ten years \citep{harris_forecasting_2018}. In contrast, datasets spanning spatial gradients capture the long-term outcome of interactions between fast processes and slower processes such as ecological and evolutionary selection, dispersal, and responses to large changes in the environment. Models based on data from spatial gradients may be a better choice for predicting long-term, steady-state outcomes of ecological change.

@@

An alternative to a weighted combination of predictions from time-series and space-for-time models is to rely on fully process-based models. If we could accurately characterize all of the processes governing a system, then a model based on that understanding should make accurate predictions at all time-horizons. For example, rather than fitting phenomenological models to our simulated time-series, we could have fit the exact process-based models that we used to generate those time-series. Assuming reasonable estimates of the parameter, those models would have accurately predicted the dynamics at all forecast horizons. Process-based models should also be more robust for making predictions outside of historically observed conditions and even beyond the conditions observed across spatial gradients, which will be especially important for making predictions in a future with increasingly novel combinations of environment and species interactions \citep{Williams2007}. Unfortunately, in most cases this approach is not currently feasible because we lack a detailed knowledge of all the complex and interacting processes influencing the dynamics of real ecological systems. Even if the general form of the models was known, estimating the high number of parameters and quantifying how they vary across ecosystems typically requires more data than is currently available even for well studied systems. As a result, models used for ecological forecasting will include at least some phenomenological components. But that does not mean that phenomenological forecast models do not benefit from process-based understanding. The message from our simulations is that different processes should be considered for different forecast time-scales, and this can be done by fitting models to different kinds of datasets. Even when process-level understanding does not enable a fully mechanistic model, it can improve the specification of phenomenological models.

@@

@@

The two simulation models in the main text describe how population size, $N(x,t)$, at location $x$ changes over time ($t$). We assume that the temperature, $K(x,t)$, at each location can vary in time and space.

@@

\noindent \textbf{Haploid Model:} Begin with a haploid model that describes the number of seeds present in a seed bank.  $N_{i,t}$ is the number of seeds of species $i$ at time $t$.  The model is
\begin{align}\begin{split}
N_{1,t+1} &= s_1 [1-g_1(K(t))]N_{1,t}+\frac{\lambda_1g_1(K(t))N_{1,t}}{1+ \alpha_{11}g_1(K(t))N_{1,t} + \alpha_{12}g_2(K(t))N_{2,t}}\\
N_{2,t+1} &= s_2 [1-g_2(K(t))]N_{2,t}+\frac{\lambda_2g_2(K(t))N_{2,t}}{1+ \alpha_{21}g_1(K(t))N_{1,t} + \alpha_{22}g_2(K(t))N_{2,t}}
\end{split}\end{align}
where $g_{i}(K(t))$ is the probability of germination, $K(t)$ is the temperature at time $t$, $s_i$ is the seed survival probability for species $i$, and $\lambda_i$ is the seed production rate per plant.  Below we refer to the $\alpha_{ij}$ as intra- and inter-specific competition coefficients.

@@

Now consider the case where the competition coefficients are non-zero ($\alpha_{i,j}\neq0$).  Including competition changes the way in which we compute $T$, $\phi_A$, and $\phi_a$.  Specifically, because the total number of seeds produced per year by each genotypes is reduced based on intra- and inter-specific competition, the total number of gamete-pairs becomes
\begin{align}\begin{split}
T &=  \frac{\lambda_{AA}N_{AA}(t)g_{AA}(K(t))}{1+ \alpha_{AA,AA}g_{AA}(K(t))N_{AA}(t) + \alpha_{AA,Aa}g_{Aa}(K(t))N_{Aa}(t)+ \alpha_{AA,aa}g_{aa}(K(t))N_{aa}(t)} \\ 
&+ \frac{\lambda_{Aa}N_{Aa}(t)g_{Aa}(K(t))}{1+ \alpha_{Aa,AA}g_{AA}(K(t))N_{AA}(t) + \alpha_{Aa,Aa}g_{Aa}(K(t))N_{Aa}(t)+ \alpha_{Aa,aa}g_{aa}(K(t))N_{aa}(t)}\\
&+\frac{\lambda_{aa}N_{aa}(t)g_{aa}(K(t))}{1+ \alpha_{aa,AA}g_{AA}(K(t))N_{AA}(t) + \alpha_{aa,Aa}g_{Aa}(K(t))N_{Aa}(t)+ \alpha_{aa,aa}g_{aa}(K(t))N_{aa}(t)}.
\label{eqn:defineT}
\end{split}\end{align}
The first line is the number of gamete-pairs produced by $AA$ individuals after accounting for the effects of competition.  The second and third lines are the numbers of gamete-pairs produced by $Aa$ and $aa$ individuals, respectively. The proportions of $A$ gametes and $a$ gametes are 
\begin{align}\begin{split}
\phi_A &= \frac{1}{T}\frac{\lambda_{AA}N_{AA}(t)g_{AA}(K(t))}{1+ \alpha_{AA,AA}g_{AA}(K(t))N_{AA}(t) + \alpha_{AA,Aa}g_{Aa}(K(t))N_{Aa}(t)+ \alpha_{AA,aa}g_{aa}(K(t))N_{aa}(t)} \\
&+ \frac{1}{2T}\frac{\lambda_{Aa}N_{Aa}(t)g_{Aa}(K(t))}{1+ \alpha_{Aa,AA}g_{AA}(K(t))N_{AA}(t) + \alpha_{Aa,Aa}g_{Aa}(K(t))N_{Aa}(t)+ \alpha_{Aa,aa}g_{aa}(K(t))N_{aa}(t)}\\
\phi_a &= 1- \phi_A
\label{eqn:definePhi}
\end{split}\end{align}
Combining all of this results in the same model as above,
\begin{align}\begin{split}
N_{AA}(t+1) &= s_{AA}[1-g_{AA}(K(t))]N_{AA}(t) + \phi_A^2T\\
N_{Aa}(t+1) &= s_{Aa}[1-g_{Aa}(K(t))]N_{Aa}(t) + 2 \phi_A\phi_aT\\
N_{aa}(t+1) &= s_{aa}[1-g_{aa}(K(t))]N_{aa}(t) + \phi_a^2T,
\end{split}\end{align}
but the definitions of $T$, $\phi_A$, and $\phi_a$ are given by equations (\ref{eqn:defineT}) and (\ref{eqn:definePhi}) .

>>>>

\author[1]{Peter B. Adler}  %\thanks{Corresponding author. Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah Email: peter.adler@usu.edu}}
\author[2,3,4]{Ethan P. White}
\author[5]{Michael H. Cortez}
\affil[1]{Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah}
\affil[2]{Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida}
\affil[3]{Informatics Institute, University of Florida, Gainesville, Florida}
\affil[4]{Biodiversity Institute, University of Florida, Gainesville, Florida}
\affil[5]{Department of Biological Science, Florida State University, Tallahasee, Florida}

@@

We made forecasts based on two phenomenological models, each representing processes operating at different time scales. One model represents the time-series or ``temporal approach.'' We correlated interannual variation in an ecological response with interannual variation in the environment at just one site. The other model represents the space-for-time substitution approach, which we call the ``spatial approach'' for brevity. We correlated the mean environment with the mean of an ecological state or rate across many sites.  We compared forecasts from both models to the simulated dynamics to determine how well the two approaches performed at different forecast horizons. We also assessed the potential for combining the information available in temporal and spatial patterns by using a weighted average of the forecasts from the temporal and spatial approaches optimized to best match the (simulated) observations. We then studied how the optimal model weights changed over time. We expected the temporal model to best predict short-term dynamics, the spatial model to best predict long-term dynamics, and that a combination of the two approaches would provide the best forecasts at transitional, intermediate time scales. The spatial, temporal and combined statistical models are described in Supporint Information (Appendix \ref{models}) and all code for both case studies is available at Github (https://github.com/pbadler/space-time-forecast).

@@

Conservation biologists and natural resource managers often need to anticipate the impact of environmental change on the abundance of 
endangered species, biological invaders, and harvested species. Although the managers may be primarily interested in just one focal species, skillful prediction might require considering interactions with many other species, greatly complicating the problem. But at what forecast horizon do altered species interactions become impossible to ignore? We explored this question using a metacommunity model developed by \cite{alexander_lags_2018} to study how community responses to increasing temperature depend on the interplay between within-site demography and competitive interactions and the movement of species across sites. The model features Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient. Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by  fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability. We provide a more detailed description of the model in SI Appendix \ref{metacomm}.

@@

During the baseline period there were strong spatial patterns across the mean temperature gradient. 
Individual species, including our focal species, showed classic, unimodal ``Whittaker" patterns of abundances across the gradient (Fig. \ref{fig:species-patterns-models}A).
These spatial patterns are the basis for our ``spatial model" of the temperature-biomass relationship for our focal species (Fig. \ref{fig:species-patterns-models}A). 
In contrast to the strong spatial patterns, population and community responses to interannual variation in temperature within sites were weak.
At our focal site in the center of the gradient, the biomass of the focal species was quite insensitive to interannnual variation in temperature, but showed strong temporal autocorrelation (Fig. \ref{fig:species-patterns-models}B). Our ``temporal model" estimates this weak, linear temperature effect, along with the strong lag effect of biomass in the previous year.

@@

We fit both a temporal and a spatial statistical model to forecast the effect of a temperature increase (Fig. \ref{fig:community-forecast-species}A) on the focal species' biomass at one location in the center of the temperature gradient. 
The predictions from the spatial and temporal models contrasted markedly, with the temporal 
model predicting a large increase in biomass and the spatial model predicting a decrease. Initially, the simulated abundances followed the increase predicted by the 
temporal model, but as faster-growing species colonized and increased in abundance at the focal site, the biomass of the focal species decreased, eventually falling below its baseline level (Fig. \ref{fig:community-forecast-species}B).

To combine the temporal and spatial model into a single forecast, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, the combined population forecast is $\omega \cdot T(N_{t-1},K_t) + (1-\omega) \cdot S(K_t) $ where $T$ is the temporal model, which depends on population size, $N$, and expected temperature, $K$, and $S$ is the spatial model, which depends only on $K$ (see SI Appendix \ref{models} for a full description of the approach).
This combined model accurately predicts the simulated dynamics across the full forecast horizon (Fig. \ref{fig:community-forecast-species}B), but 
the reason it works so well is that the weights were determined by fitting directly to the data. Unlike our spatial and temporal model forecasts, 
we did not generate out-of-sample predictions from the combined model; it merely provides 
a convenient way to quantify how rapidly dynamics shift from being dominated by interannual 
variation captured in the temporal model (time $t=0$ to $t \approx 1250$ in Fig. \ref{fig:community-forecast-species}B) to being dominated by the steady-state equilibrium captured by the spatial model (time $t \geq 2500$).
A true, combined forecast would require a method to determine the model weights \emph{a priori}.
The most rapid shifts in the model weights occurred during the period when warm-adapted, faster growing species were increasing most rapidly
in abundance (Fig. \ref{fig:community-forecast-species}C).

@@

The compositional turnover affecting our focal species also influences total biomass, linking community and ecosystem dynamics. We repeated our focal species analysis for total community biomass, and the results were similar: the temporal model initially made the best forecasts immediately following the onset of the temperature increase, but as the identity and abundances of species at the study site changed, the model weights rapidly shifted to the spatial model (SI Figs. \ref{fig:community-models-total} and \ref{fig:community-forecasts-total}).

@@

Evolutionary adaptation is a key uncertainty in predicting how environmental change will impact a focal population at a given location \citep{Hoffmann2011}. Like the shifts in species composition illustrated in the previous example, shifts in genotype frequencies can also influence dynamics and forecasts at different time scales. Although shifts in genotype frequencies at the population level are analogous to changes in species composition at the community level, the mechanisms are distinct: heterozygosity and genetic recombination have no analogue at the community level. We demonstrate how these processes influence short and long-term forecasts with a standard eco-evolutionary model for a hypothetical annual plant population in which fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A). Our model describes how the local density of each genotype changes between years,  which depends on temperature and genotype densities in the previous year. Transient temporal dynamics are computed directly from the model; these dynamics define our temporal forecast. To create ``spatial data'', we simulated the equilibrium density of each genotype under different mean temperatures. The pattern of equilibrium densities across a gradient in mean annual temperature defines our spatial forecast: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B). The full model description is provided in SI Appendix \ref{eco-evo}.

@@

We designed our simulation studies to illustrate how the change in model performance with increasing forecast horizon reflects differences in the types and scales of processes captured by spatial and temporal data sets. Local time-series data capture demographic processes, lagged effects, and responses of current assemblages to small changes in environmental conditions. In addition, the state of the system in the near future depends heavily on the current state. Since short-term forecasts do not typically require extrapolating into novel conditions, a model based on the historical range of variation which incorporates lags and accurate initial conditions is likely to be the most generally useful approach for forecasting. The hypothesis that time-series models will be most effective for near-term forecasts is supported by recent empirical analyses of biodiversity forecasts at time scales from one to ten years \citep{harris_forecasting_2018}. In contrast, datasets spanning spatial gradients should capture the long-term outcome of interactions between fast processes and slower processes such as ecological and evolutionary selection, dispersal, and responses to large changes in the environment. Models based on data from spatial gradients may be a better choice for predicting long-term, steady-state outcomes of ecological change \citep{Blois2013}.

@@

An alternative to a weighted combination of predictions from time-series and space-for-time models is to rely on fully process-based models. If we could accurately characterize all of the processes governing a system, then a model based on that understanding should make accurate predictions at all time-horizons. For example, rather than fitting phenomenological models to our simulated time-series, we could have fit the exact process-based models that we used to generate those time-series. Assuming reasonable estimates of the parameter, those models would have accurately predicted the dynamics at all forecast horizons. Process-based models should also be more robust for making predictions outside of historically observed conditions and even beyond the conditions observed across spatial gradients, which will be especially important for making predictions in a future with increasingly novel combinations of environment and species interactions \citep{Williams2007}. Unfortunately, in most cases this approach is not currently feasible because we lack a detailed knowledge of all the complex and interacting processes influencing the dynamics of real ecological systems. Even if the general form of the models was known, estimating the high number of parameters and quantifying how they vary across ecosystems typically requires more data than is currently available even for well studied systems. 
% mention difficulty of dealing with error propogation?
As a result, models used for ecological forecasting will include at least some phenomenological components. But that does not mean that phenomenological forecast models do not benefit from process-based understanding. The message from our simulations is that different processes should be considered for different forecast time-scales, and this can be done by fitting models to different kinds of datasets. Even when process-level understanding does not enable a fully mechanistic model, it can improve the specification of phenomenological models.

@@

\subsubsection*{Acknowledgements}

We thank Heather Lynch and Juan Manuel Morales for comments that improved early drafts of the paper. PBA was supported by the Utah Agriculture Experiment Station. EPW was supported by the Gordon and Betty Moore Foundation’s Data-Driven Discovery Initiative through Grant GBMF4563.

@@

The two simulation models in the main text describe how population size, $N(x,t)$, at location $x$ changes over time ($t$). We assume that the temperature, $K(x,t)$, at each location can vary in time and space. To forecast the dynamics generated by these simulations models, we fit a series of statistical models.

@@

\noindent \textbf{Haploid Model:} Begin with a haploid model that describes the number of seeds present in a seed bank.  $N_{i,t}$ is the number of seeds of species $i$ at time $t$.  The model is
\begin{align}\begin{split}
N_{1,t+1} &= s_1 [1-g_1(K(t))]N_{1,t}+\frac{\lambda_1g_1(K(t))N_{1,t}}{1+ \alpha_{11}g_1(K(t))N_{1,t} + \alpha_{12}g_2(K(t))N_{2,t}}\\
N_{2,t+1} &= s_2 [1-g_2(K(t))]N_{2,t}+\frac{\lambda_2g_2(K(t))N_{2,t}}{1+ \alpha_{21}g_1(K(t))N_{1,t} + \alpha_{22}g_2(K(t))N_{2,t}}
\end{split}\end{align}
where $g_{i}(K(t))$ is the probability of germination, $K(t)$ is the temperature at time $t$, $s_i$ is the seed survival probability for species $i$, and $\lambda_i$ is the seed production rate per plant.  Below we refer to the $\alpha_{ij}$ as intra- and inter-genotype competition coefficients.

@@

Now consider the case where the competition coefficients are non-zero ($\alpha_{i,j}\neq0$).  Including competition changes the way in which we compute $T$, $\phi_A$, and $\phi_a$.  Specifically, because the total number of seeds produced per year by each genotypes is reduced based on intra- and inter-genotype competition, the total number of gamete-pairs becomes
\begin{align}\begin{split}
T &=  \frac{\lambda_{AA}N_{AA}(t)g_{AA}(K(t))}{1+ \alpha_{AA,AA}g_{AA}(K(t))N_{AA}(t) + \alpha_{AA,Aa}g_{Aa}(K(t))N_{Aa}(t)+ \alpha_{AA,aa}g_{aa}(K(t))N_{aa}(t)} \\ 
&+ \frac{\lambda_{Aa}N_{Aa}(t)g_{Aa}(K(t))}{1+ \alpha_{Aa,AA}g_{AA}(K(t))N_{AA}(t) + \alpha_{Aa,Aa}g_{Aa}(K(t))N_{Aa}(t)+ \alpha_{Aa,aa}g_{aa}(K(t))N_{aa}(t)}\\
&+\frac{\lambda_{aa}N_{aa}(t)g_{aa}(K(t))}{1+ \alpha_{aa,AA}g_{AA}(K(t))N_{AA}(t) + \alpha_{aa,Aa}g_{Aa}(K(t))N_{Aa}(t)+ \alpha_{aa,aa}g_{aa}(K(t))N_{aa}(t)}.
\label{eqn:defineT}
\end{split}\end{align}
The first line is the number of gamete-pairs produced by $AA$ individuals after accounting for the effects of competition.  The second and third lines are the numbers of gamete-pairs produced by $Aa$ and $aa$ individuals, respectively. The proportions of $A$ gametes and $a$ gametes are 
\begin{align}\begin{split}
\phi_A &= \frac{1}{T}\frac{\lambda_{AA}N_{AA}(t)g_{AA}(K(t))}{1+ \alpha_{AA,AA}g_{AA}(K(t))N_{AA}(t) + \alpha_{AA,Aa}g_{Aa}(K(t))N_{Aa}(t)+ \alpha_{AA,aa}g_{aa}(K(t))N_{aa}(t)} \\
&+ \frac{1}{2T}\frac{\lambda_{Aa}N_{Aa}(t)g_{Aa}(K(t))}{1+ \alpha_{Aa,AA}g_{AA}(K(t))N_{AA}(t) + \alpha_{Aa,Aa}g_{Aa}(K(t))N_{Aa}(t)+ \alpha_{Aa,aa}g_{aa}(K(t))N_{aa}(t)}\\
\phi_a &= 1- \phi_A
\label{eqn:definePhi}
\end{split}\end{align}
Combining all of this results in the same model as above,
\begin{align}\begin{split}
N_{AA}(t+1) &= s_{AA}[1-g_{AA}(K(t))]N_{AA}(t) + \phi_A^2T\\
N_{Aa}(t+1) &= s_{Aa}[1-g_{Aa}(K(t))]N_{Aa}(t) + 2 \phi_A\phi_aT\\
N_{aa}(t+1) &= s_{aa}[1-g_{aa}(K(t))]N_{aa}(t) + \phi_a^2T,
\end{split}\end{align}
but the definitions of $T$, $\phi_A$, and $\phi_a$ are given by equations (\ref{eqn:defineT}) and (\ref{eqn:definePhi}) .
